---
title: "fMRI Preprocessing using SPM 12"
author: "John Muschelli"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fMRI Preprocessing using SPM 12}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r knit-setup, echo=FALSE, results='hide', eval=TRUE, cache = FALSE, warning = FALSE, message = FALSE}
library(spm12r)
```


## Overview of `spm12r` functions

Requires MATLAB (installs SPM to R library)

- `spm12_slice_timing` - slices are not taken instantaneously
- `spm12_realign` - people move
- `spm12_coregister` - structural imaging is higher resolution
- `spm12_segment` - where's the gray matter?
- `spm12_normalize` - brains look better in MNI
- `spm12_smooth` - turn that noise down


```{r makefiles}
library(kirby21.t1)
library(kirby21.fmri)
functional = get_fmri_filenames(ids = 113, visit = 1)
anatomical = get_t1_filenames(ids = 113, visit = 1)
files = c(anatomical = anatomical,
          functional = functional)
files
```

```{r}
library(neurobase)
tr = 2 # seconds
DROP = 10 # 20 seconds for stabilization

fmri = readnii(files["functional"])

times = (DROP + 1):ntim(fmri)
run_fmri = copyNIfTIHeader(fmri, fmri[,,,times], drop = TRUE)
```

```{r, echo = FALSE}
rm(list = "fmri")
```


```{r nb_version, echo = FALSE, results="hide"}
library(neurobase)
ip = installed.packages()
ver = ip["neurobase", "Version"]
ver = compareVersion(ver, "1.13")
if (ver < 0) {
  remap_filename = function(
    x, 
    prefix = "", 
    suffix = "") {
    
    ext = parse_img_ext(x)
    ext = paste0(".", ext)
    tx = tolower(x)
    if (grepl("gz$", tx)) {
      ext = paste0(ext, ".gz")
    }
    dn = dirname(x)
    stub = nii.stub(x, bn = TRUE)
    
    file.path(
      dn,
      paste0(
        prefix, 
        stub,
        suffix,
        ext)
    )
  }
}
```

```{r have_matlab}
library(matlabr)
have_matlab()
```

```{r realign}
library(spm12r)
####################################
# Realignment
####################################
if (have_matlab()) {
  realigned = spm12_realign( 
	filename = run_fmri, 
	register_to = "mean",
	reslice = "mean",
	clean = FALSE
	)
  print(realigned)
}
```

```{r rp_file}
####################################
# Read in Motion data
####################################
if (have_matlab()) {
  rpfile = realigned[['rp']]
  rp = read.table(file = rpfile, header = FALSE)
  colnames(rp) = c("x", "y", "z", 
  	"roll", "pitch", "yaw")
  rp = as.matrix(rp)
  print(head(rp))
}
```

```{r slice_time}
####################################
# Slice Timing Correction
####################################
nslices = oro.nifti::nsli(run_fmri)
slice_order = 1:nslices
ref_slice = slice_order[median(seq(nslices))]
ta = tr - tr/nslices
n_time_points = ntim(run_fmri)
if (have_matlab()) {
  aimg = spm12_slice_timing(
  	filename = realigned[['outfiles']],
  	nslices = nslices,
  	tr = tr, 
  	slice_order = slice_order,
  	ta = ta, 
  	ref_slice = ref_slice,
  	prefix = "a", 
  	clean = FALSE, 
  	retimg = FALSE)
  mean_img = realigned[["mean"]]
}
```  

```{r acpc}
if (have_matlab()) {
  acpc_reorient(
    infiles = c(mean_img, aimg),
    modality = "T1")
}
```

## Direct Normalization

```{r direct_norm}
if (have_matlab()) {
  bbox = matrix(
  		c(-90, -126, -72, 
  		90, 90, 108), 
  		nrow = 2, byrow = TRUE)
  print(bbox)
  direct_norm = spm12_normalize(
  	filename = mean_img,
  	other.files = c(mean_img, aimg),
  	bounding_box = bbox,
  	clean = FALSE
  	)
  dnorm_files = direct_norm$outfiles
  dnorm_mean_img = readnii(dnorm_files[1])
}
```


# Indirect Normalization

## Anatomical MRI Coregistration to Mean fMRI
```{r coreg}
if (have_matlab()) {
  anatomical = files["anatomical"]
  anat_img = checknii(anatomical)
  print(anat_img)
  acpc_reorient(
    infiles = anat_img,
    modality = "T1")

  coreg = spm12_coregister(
  	fixed = mean_img,
  	moving = anat_img,
  	prefix = "r")
  
  coreg_anat = remap_filename(anat_img, prefix = "r")
}
```

## Anatomical MRI Segmentation (and Spatial Normalize Estimation)

```{r seg}
if (have_matlab()) {
  seg_res = spm12_segment(
  	filename = coreg_anat,
  	retimg = FALSE)
  print(seg_res)
}
```

## Applying Spatial Normalize Transformation

```{r norm_write}
bbox = matrix(
  c(-90, -126, -72, 
    90, 90, 108), 
  nrow = 2, byrow = TRUE)
if (have_matlab()) {
  norm = spm12_normalize_write(
  	deformation = seg_res$deformation,
  	other.files = c(coreg_anat, mean_img, aimg),
  	bounding_box = bbox,
  	retimg = FALSE, 
  	clean = FALSE)
  print(norm)
  norm_data = norm$outfiles
  names(norm_data) = c("anat", "mean", "fmri")
  norm_mean_img = readnii(norm_data["mean"])
}
```



```{r check_norm}
if (have_matlab()) {
  template_path = file.path(spm_dir(), 
                            "canonical", "avg152T1.nii")
  
  template = readnii(template_path)
  double_ortho(template, norm_mean_img)
  double_ortho(template, dnorm_mean_img)
  double_ortho(norm_mean_img, dnorm_mean_img)
}
```

```{r smooth}
if (have_matlab()) {
  smoothed = spm12_smooth(
  	filename = norm_data["fmri"],
  	prefix = "s",
  	retimg = FALSE
  	)
  smoothed_data = smoothed$outfiles
}

```

```{r smooth_mean}
if (have_matlab()) {
  smoothed_mean = spm12_smooth(
  	filename = norm_data["mean"],
  	prefix = "s",
  	retimg = FALSE
  	)  
  smoothed_mean_data = smoothed_mean$outfiles
}
```

```{r plot_smoothed}
smooth_img = readnii(smoothed_data, drop_dim = FALSE)
```

## Bibliography


